A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-10-10 doc/large_strain.tex
spaces in equatinos
various improvements to the text

The performance of finite element (FEM) solvers on modern computer architectures is typically memory bound.
The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
In order to improve the performance of the iterative solver, so-called matrix-free methods are
widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}. Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016}.

To the best of our knowledge, the matrix-free methods are not yet widely adopted within the solid mechanics community.
We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a master thesis of a student of two of the authors \cite{Mentler2017}.
A possible reason is that tangent operators (coming from linearization of the nonlinear balance equations) are
more elaborate as compared to the fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case. Another reason could be the smoothness of the underlying solution. FEM with linear or quadratic elements (to avoid locking) are often adopted in solid mechanics community. However we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore we believe that for large scale computations in solid mechanics it is crucial to adopt matrix-free approaches and develop efficient multigrid solvers.

In this work, we advance the application of matrix-free approaches to problems in solid mechanics.
We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
In order to improve the convergence of iterative solvers, we also propose a way to construct level tangent operators
and employ them to define a geometric multigrid preconditioner.
Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.

@@

For conservative systems without the volumetric load the total potential energy functional $\mcl E$ is introduced as
\begin{align}
\mcl E =
\int_{\mcl B_0} \mcl \psi(\gz F) \d V \,
- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \d S \, ,
\end{align}
where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.

The principal of stationary potential energy requires that at the equilibriums the directional derivative with respect to the displacement $\gz u$
\begin{align}
\delta \mcl E =
\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
\label{eq:stationary}
\end{align}
vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
%
This leads to the following scalar non-linear equation
\begin{align}
F(\gz u, \delta \gz u) =
\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \d V \,
-
\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \d S
= 0 \, ,
\end{align}
where $\gz P := \partial \mcl \psi / \partial \gz F$ is the first Piola-Kirchhoff stress tensor.
The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
\begin{align}
\gz P : \Grad \, \delta \gz u =
(\gz \tau \cdot \gz F^{-T}) : \Grad \, \delta \gz u =
\gz \tau : (\Grad \, \delta \gz u \cdot \gz F^{-1}) =
\gz \tau : \grad \, \delta \gz u \, ,
\end{align}
and therefore
\begin{align}
  F(\gz u, \delta \gz u) =
  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \d S
  = 0 \, ,
\label{eq:weak_form}
\end{align}
where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
Note that $\gz \tau$ is the Piola transformation of the second Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.

@@

In order to apply the Newton method, linearization of \eqref{eq:weak_form} around a given solution field $\overline{\gz u}$ is required. This is given by
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{align}
\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
=
\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \d V
+
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V
\end{align}
This can be simplified \cite{Wriggers2008} to
\begin{align}
\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  =
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V
\label{eq:tangent}
\end{align}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$ and the fourth order tangent tensor is push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is called material part of the tangent operator as it is related to the linearization of the Kirchhoff stress. Whereas the second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from linearization of the deformation gradient $\gz F$.

@@

We now introduce a FE triangulation $\mathcal{T}^h$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements of a fixed polynomial degree. % : $V^h \subset H^1 (\mcl B_0)$.
The displacement fields are given in a
a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support):
\begin{alignat}{2}
       \gz u^h &=:  \sum_{i \in \mcl I_\varphi} u^i \gz N_i (\gz X) \quad \quad \quad
\delta \gz u^h &&=: \sum_{i \in \mcl I_\varphi} \delta u^i \gz N_i (\gz X) \,,
\end{alignat}
where superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
For the sake of this study $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.

The Newton-Raphson solution approach is adopted for the here considered nonlinear problem.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is to be obtained as a solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent})
\begin{align}
  K_{ij} \Delta u_j &= - F_i \\
  K_{ij} &=
  \int_{\mcl B_0} \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  \d V
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \d S
\end{align}
Here $\gz K$ is the discrete tangent operator and $\gz F$ is the discrete gradient.

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically the approach for finding a discrete solution to a linearized partial differential equation can be described as follows:
First, ansatz spaces and a suitable bilinear form is defined. Then, a matrix and force vector corresponding to the discretization are assembled.
Finally, a direct or iterative solver is used to solve the linear system for the discretized solution.
On modern computer architectures however, it turns out that loading data is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations in the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix vector products. More precisely, the linear system corresponding to a bilinear form
$a\colon V_h\times V_h\to\mathbb{R}$ and right-hand side $f\colon V_h\to\mathbb{R}$ defined with respect to the ansatz space $V_h$ can be represented by
\begin{align*}
 &&a(v,u) &= f(v) && \forall v\in V_h\\
 \Longleftrightarrow &&a(\varphi_i, u) &= f(\varphi_i) && \forall \varphi_i \\
 \Longleftrightarrow &&\sum_j a(\varphi_i,\varphi_j) u_j &= f(\varphi_i) && \forall \varphi_i \\
 \Longleftrightarrow &&A x &= b
\end{align*}
where $A_{i,j}=a(\varphi_i, \varphi_j)$ and $b_i= f(\varphi_i)$.
Thus, the matrix-vector product $Ax$ can be expressed by
\begin{align*}
 (Ax)_i &= \sum_j a(\varphi_i,\varphi_j) x_j \\
        &= \sum_K\sum_q \sum_j \tilde{a}_K(\varphi_i,\varphi_j)(x_q) x_j w_q
\end{align*}
where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)$ is a representation for the evaluation of the bilinear in one quadrature point and a quadrature rule is used.
For the sake of demonstration, let us assume $A$ represents a mass matrix. In this case, it holds
\begin{align*}
 \tilde{a}_K(\varphi_i,\varphi_j)(x_q) = \varphi_i(x_q)\varphi_j(x_q).
\end{align*}
Further assuming that the number of the one-dimensional quadrature points $n$ matches the degree of the ansatz space plus one, the evaluation of all the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in case we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$
and this is the same as for a regular matrix-vector where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, we can actually do better
than that. Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
\begin{lstlisting}
template<int k, int q>
void evaluate_dof_vector_in_all_quadrature_points
  (const double u[k][k][k],
   const double phi[k][q],
   double return_value[q][q][q])
{
  double tmp1[k][k][q] {};
  for (int i1=0; i1<k; ++i1)
    for (int i2=0; i2<k; ++i2)
      for (int q3=0; q3<q; ++q3)
        for (int i3=0; i3<k; ++i3)
          tmp1[i1][i2][q3] += u[i1][i2][i3]*phi[i3][q3];

@@

  for (int q2=0; q2<q; ++q2)
    for (int q3=0; q3<q; ++q3)
      for (int q1=0; q1<q; ++q1)
        for (int i1=0; i1<k; ++i1)
  return_value[q1][q2][q3] += tmp2[i1][q2][q3]*phi[i1][q1];
}
\end{lstlisting}
in three space dimensions. In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-base especially for 3D and high polynomial degree.
More information on techniques to increase the computational density can be found in \cite{kronbichler12,vos10}.
We briefly note that the SIMD vectorization is applied at the element level, that is the operator is applied simultaneously on several several elements (called ``block'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.

Given that the matrix-free operators are applied on-the-fly, it's crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below we propose three algorithms to implement tangent operator of finite-strain linear elasticity using matrix-free approach, that employ different caching strategies.
%
In order to obtain gradients with respect to the deformed configuration we adopted \textit{MappingQEulerian} class from \texttt{deal.II} library.
In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, this class adds mapping to the deformed configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, when performing contraction with gradient of shape functions in current (deformed) configuration, the integration is also done over the deformed configuration. Therefore in order to arrive at the integral in undeformed configuration, additionally we have to divide by Jacobian $J$ of the mapping at a given quadrature point.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$, current FE solution $\gz u$,
  cacheed $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $\gz F = \gz I + \gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
  }
  \caption{Matrix-free tangent cell operator: scalar cache}
  \label{alg:mf_scalar}
\end{algorithm}

Algorithm \ref{alg:mf_scalar} caches a single scalar which involves logarithm of Jacobian -- a computationally demanding operation compared to addition and multiplications. As evident from Section \ref{sec:theory}, in order to apply the operator we need to re-evaluate Kirchhoff stress at each quadrature point. This in turn requires evaluation of deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point}
  \Output{action of the FE tangent on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\boldsymbol{\mathcal{G}}\gz g = \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s = c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
  }
  \caption{Matrix-free tangent cell operator, variant 1: cache second order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate refential quantities like $\gz F$ at runtime.}
  \label{alg:mf_tensor2}
\end{algorithm}

In order to avoid re-calculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. This also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration. This algorithm utilize the chosen constitutive relationship and therefore the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ is cheap to evaluate, see \eqref{eq:simplified_action}. Here we also cache two scalars that depend on the Jacobian $J$.

>>>>

The performance of finite element (FEM) solvers on modern computer architectures is typically memory bound.
The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
In order to improve the performance of the iterative solver, so-called matrix-free methods are
widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}. Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016}.
{\color{red}Daniel: please add more links if you want, maybe GPU related}.

To the best of our knowledge, the matrix-free methods are not yet widely adopted within the solid mechanics community.
We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a master thesis of a student of two of the authors \cite{Mentler2017}.
A possible reason is that tangent operators (coming from linearization of the nonlinear balance equations) are
more elaborate as compared to the fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case. Another reason could be the smoothness of the underlying solution. FEM with linear or quadratic elements, that avoid locking, are often adopted in solid mechanics community. Higher order elements are rarely employed. However we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore we believe that for large scale computations in solid mechanics it is crucial to adopt matrix-free approaches and develop efficient multigrid solvers.

In this work, we advance the application of matrix-free approaches to problems in solid mechanics.
We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
In order to improve the convergence of iterative solvers, we also propose a way to construct level matrix-free tangent operators
and employ them to define a geometric multigrid preconditioner.
Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.

@@

For conservative systems without the volumetric load the total potential energy functional $\mcl E$ is introduced as
\begin{align}
\mcl E =
\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
\end{align}
where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.

The principal of stationary potential energy requires that at the equilibriums the directional derivative with respect to the displacement $\gz u$
\begin{align}
\delta \mcl E =
\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
\label{eq:stationary}
\end{align}
vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
%
This leads to the following scalar non-linear equation
\begin{align}
F(\gz u, \delta \gz u) =
\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \, \d V \,
-
\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
= 0 \, ,
\end{align}
where $\gz P := \partial \mcl \psi / \partial \gz F$ is the first Piola-Kirchhoff stress tensor.
The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
\begin{align}
\gz P : \Grad \, \delta \gz u =
(\gz \tau \cdot \gz F^{-T}) : \Grad \, \delta \gz u =
\gz \tau : (\Grad \, \delta \gz u \cdot \gz F^{-1}) =
\gz \tau : \grad \, \delta \gz u \, ,
\end{align}
and therefore
\begin{align}
  F(\gz u, \delta \gz u) =
  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
  = 0 \, ,
\label{eq:weak_form}
\end{align}
where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
Note that $\gz \tau$ is the Piola transformation of the second Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.

@@

In order to apply the Newton method, linearization of \eqref{eq:weak_form} around a given solution field $\overline{\gz u}$ is required. This is given by
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{align}
\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
=
\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
+
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V
\end{align}
This can be simplified \cite{Wriggers2008} to
\begin{align}
\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  =
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V
\label{eq:tangent}
\end{align}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$ and the fourth order tangent tensor is push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is called material part of the tangent operator as it is related to the linearization of the Kirchhoff stress. Whereas the second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from linearization of the deformation gradient $\gz F$.

@@

We now introduce a FE triangulation $\mathcal{T}^h$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements of a fixed polynomial degree. % : $V^h \subset H^1 (\mcl B_0)$.
The displacement fields are given in a
a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
\begin{alignat}{2}
       \gz u^h &=:  \sum_{i \in \mcl I_\varphi} u^i \gz N_i (\gz X) \quad \quad \quad
\delta \gz u^h &&=: \sum_{i \in \mcl I_\varphi} \delta u^i \gz N_i (\gz X) \,,
\end{alignat}
where superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
For the sake of this study $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.

The Newton-Raphson solution approach is adopted for the here considered nonlinear problem.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is to be obtained as a solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent})
\begin{align}
  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
  A_{ij} &=
  \int_{\mcl B_0}
  \underbrace{
  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
  +
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  }_{\tilde{a}(\gz N_i, \gz N_j)}
  \d V
  \label{eq:algebraic_tangent}
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S
\end{align}
Here $\gz A$ is the discrete tangent operator, $\gz F$ is the discrete gradient
and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the bilinear form for the trial solution field $\overline{\gz u^h}$.

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically in order to solve \eqref{eq:linear_system}, the matrix $\gz K$ and the force vector $\gz F$ corresponding to the discretization are assembled and
a direct or iterative solver is used to solve the linear system.
On modern computer architectures however, it turns out that loading data is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations in the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix vector products.
%
The matrix-vector product $A x$ can be expressed by
\begin{align*}
 (Ax)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &= \sum_K\sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j w_q
\end{align*}
where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)(\gz x_q)$ is a representation for the evaluation of the bilinear in one quadrature point and a quadrature rule is used.
For the sake of demonstration, let us assume $A$ represents a mass matrix. In this case, it holds
\begin{align*}
 \tilde{a}_K(N_i,N_j)(\gz x_q) = N_i(\gz x_q)N_j(\gz x_q).
\end{align*}
Further assuming that the number of the one-dimensional quadrature points $n$ matches the degree of the ansatz space plus one, the evaluation of all the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in case we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$
and this is the same as for a regular matrix-vector where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, we can actually do better
than that. Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
\begin{lstlisting}
template<int k, int q>
void evaluate_dof_vector_in_all_quadrature_points
  (const double u[k][k][k],
   const double phi[k][q],
   double return_value[q][q][q])
{
  double tmp1[k][k][q] {};
  for (int i1=0; i1<k; ++i1)
    for (int i2=0; i2<k; ++i2)
      for (int q3=0; q3<q; ++q3)
        for (int i3=0; i3<k; ++i3)
          tmp1[i1][i2][q3] += u[i1][i2][i3]*phi[i3][q3];

@@

  for (int q2=0; q2<q; ++q2)
    for (int q3=0; q3<q; ++q3)
      for (int q1=0; q1<q; ++q1)
        for (int i1=0; i1<k; ++i1)
  return_value[q1][q2][q3] += tmp2[i1][q2][q3]*phi[i1][q1];
}
\end{lstlisting}
in three space dimensions. In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-base especially for 3D and high polynomial degree.
More information on techniques to increase the computational density can be found in \cite{kronbichler12,vos10}.
We briefly note that the SIMD vectorization is applied at the element level, that is the operator is applied simultaneously on several elements (called ``block'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.

Given that the matrix-free operators are applied on-the-fly, it's crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using matrix-free approach.
%
In order to obtain gradients with respect to the deformed configuration we adopted \textit{MappingQEulerian} class from \texttt{deal.II} library.
In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, given a displacement field this class adds mapping to the deformed configuration.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, when performing contraction with gradient of shape functions in current (deformed) configuration, the integration is also done over the deformed configuration. Therefore in order to arrive at the integral in undeformed configuration, additionally we have to divide by Jacobian $J$ of the mapping at a given quadrature point.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$, current FE solution $\gz u$,
  cacheed $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $\gz F = \gz I + \gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
  }
  \caption{Matrix-free tangent cell operator: scalar cache}
  \label{alg:mf_scalar}
\end{algorithm}

Algorithm \ref{alg:mf_scalar} caches a single scalar which involves logarithm of Jacobian -- a computationally demanding operation compared to additions and multiplications. As evident from Section \ref{sec:theory}, in order to apply the operator we need to re-evaluate Kirchhoff stress at each quadrature point. This in turn requires evaluation of deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\boldsymbol{\mathcal{G}}\gz g = \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s = c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
  }
  \caption{Matrix-free tangent cell operator, variant 1: cache second order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate refential quantities like $\gz F$ at runtime.}
  \label{alg:mf_tensor2}
\end{algorithm}

In order to avoid re-calculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration. This algorithm utilizes the chosen constitutive relationship and therefore the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ is cheap to evaluate, see \eqref{eq:simplified_action}. To that end we cache two scalars that depend on the Jacobian $J$.
